Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase Fix Time?

### 2015-03-17 phaseDelay.tex
Updates from ShareLaTeX

\conferenceinfo{FSE}{'15 Bergamo, Italy}
\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase Defect Fix Time?}
\numberofauthors{5}
\author{
\alignauthor 
Tim Menzies \\
       \affaddr{Dept. of Computer Science}\\
       \affaddr{North Carolina State U, USA}\\
       \email{tim.menzies@gmail.com}
\alignauthor 
William Nichols \\
        \affaddr{Software Engineering Institute}\\
        \affaddr{Carnegie Mellon U, USA}
        \email{wrn@sei.cmu.edu} 
\alignauthor 
Lucas Layman \\
       \affaddr{Fraunhofer CESE} \\
       \affaddr{College Park, USA}\\ 
       \email{llayman@fc-md.umd.edu}
\and
\alignauthor 
Forrest Shull \\
        \affaddr{Software Engineering Institute}\\
        \affaddr{Carnegie Mellon U, USA}
        \email{fjshull@sei.cmu.edu} 
\alignauthor 
Carter Pape \\
       \affaddr{Dept. of Computer Science}\\
       \affaddr{North Carolina State U, USA}\\
       \email{carterpape@gmail.com}
} 
\maketitle
\begin{abstract}
Does
repair time increase dramatically
the longer a defect persists in a system?
This widely held belief is used to justify 
major investments in  software development
 processes and methods.

This paper shows that this belief is maintained
quite strongly in the industrial software engineering
community (but, in the academic community, a little less so).
Yet based on a sample of 
171 software projects from around the world from 
and completed between 2006 and 2014, we argue that this belief is mostly 
incorrect. Specifically: the median fix time spent in phases
devoted to resolving issues does {\em not} increase
the longer the issue remains in the system.

This result raises the question: how do widely-held beliefs like phase delay survive so long  without  critical reassessment?  As a community, we need to reflect more on research practices and on how (and when) we reflect on the prominent hypotheses of our field.
\end{abstract}

@@

\section{Introduction}
In 2013-2014, the time of 
eleven  million programmers~\cite{pettey14} and
half a trillion dollars~\cite{avram14} was spent on information technology.
Such a large and growing effort should be managed and optimized via  well-researched conclusions.  
It is standard practice
in other fields such as medicine,
to continually revisit old conclusions~\cite{prasad13}. Software engineering should do the same lest
ill-founded guidance exert an undue influence on how we build software.

Accordingly, this paper revisits
the {\em phase delay effect}
i.e.,  the rule of thumb stating that the time required to resolve an issue in a software project increases dramatically 
the longer it remains in the system. 
Phase delay has long been
a    motivator for major changes to
SE practices. Basili \& Boehm comment that since the 1980s,  phase delay:
\begin{quote}
...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes~\cite{boehm01}.
\end{quote}
Basili and Boehm are not alone in this assessment of the significance of the phase delay effect. Our surveys of industrial practitioners (shown in \S\ref{survey}) reveal the phase delay effect to be
one of the most strongly-held beliefs amongst commercial software engineers.
Previously~\cite{me08a}, we have constructed return-on-investment models
for a common practice amongst U.S. government contractors: 
independent verification \& validation. IVV has been criticized as
ineffective since IVV engineers, many of whom are unfamiliar with the details of the project, are required to pass judgement on the code.
Hence,  the IVV team may find the same or fewer bugs
as the developers.
On the other hand, IVV has been defended
as follows:  when the IVV team finds bugs earlier than the developers then, due to the phase delay effect, this provides overall cost savings to the project. Note that the strength of this argument diminishes
if fixing later-phase issues is not significantly more expensive than fixing them earlier.

% Also, as discussed later in this paper, concerns over phase delay were a major driver in the development of agile software processes.

In this paper, we investigate several hypotheses:

\bi
    \item $H_1$: The cost-to-fix of defects monotonically increases across phases.
    \item $H_2$: The cost-to-fix of defects monotonically increases as the phase delay increases.
\ei
As we explored the above, we found a related hypothesis was believed by both professional software engineers and FSE conference reviewers:
\bi
 \item $H_3$: Requirements errors are the most expensive to fix.
\ei

The rest of the paper investigates these  hypotheses. First, we conduct a survey of practitioners and researchers to gauge their belief in the phase delay effect. Second, we  analyze 171 software  projects developed in the period 2006 to 2014.  To the best of our knowledge, this is the largest study on phase delay yet seen in the literature.

% The conclusions of this paper is that it is erroneous to believe in phase delay.
% When we looked for supporting evidence for that effect, we found that:
% \be 
% \item The evidence for that effect in the historical literature is very weak (to say the least);
% \item  In  a large sample of   contemporary software projects, there is no evidence for phase delay.
%  \ee
% The rest of this paper present evidence for the above two points.

% The first point is  documented via  a literature survey and the second point is demonstrated
% via an analysis of

@@

@@

Belief in general
principles of software
engineering are common to both research and practice. Professional societies assume such generalities exist when they offer
 lists of supposedly general ``best practices'' such as
the IEEE 1012 standard for software verification~\cite{1012}.  Endres \& Rombach's comment that 
\begin{quote}
Based on repeated and consistent observations, key lessons of these fields can now be formulated into rules or even laws, providing initial building blocks towards a theoretical foundation that is essential for further research, for teaching and for the practice of software development~\cite{endres03,rombach11}
\end{quote}
 Endres \& Rombach offer dozens of lessons of software engineering~\cite{rombach11}.
 Many other 
widely-cited researchers  do the same; e.g.
Glass~\cite{glass02}; Jones~\cite{jones07}; Boehm~\cite{boehm00b}.
Budgen \& Kitchenham seek to reorganize SE research using
general
conclusions drawn from a larger number of studies~\cite{kitch04,budgen09}.

Nevertheless, there are many empirical  findings that 
cast doubt on the existence of  ``key lessons  that can be formulated
as general laws of SE'':
\bi
\item
Turhan~\cite{me12d} lists 28 studies with contradictory conclusions
on the relation of OO measures to defects.  Those results
 directly  contradict some of the laws listed by 
Endres \& Rombach~\cite{endres03}.

@@

The cognitive issues that complicate medicine are also found in software engineering.
According to Passos et al.~\cite{passos11}, commercial developers
are all too willing to believe in general
truisms.
They  warn that developers
usually assume that the truisms they learn from a few past
projects are general to 
all their future projects:
\begin{quote}\label{q:pass}
...the past experiences were taken into account without 
much consideration for their context~\cite{passos11}.  
\end{quote}
The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} support the findings if Passos et al. They report that 
  supposed software engineering    ``experts'' rarely use lessons
  from past projects to improve their future reasoning~\cite{jorgensen09}. 
 They note that
when the experts
  fail to revise their beliefs, this leads to poor
 conclusions and software projects  (see examples in~\cite{jorgensen09}).
%% This leads to  detrimental
%% situations where groups or 
%% individuals working on the same project have conflicting beliefs, which they never debate
%or refine or unify.

@@

@@

@@

\section{A Survey on Phase Delay and Cost-to-fix Belief}
\label{survey}

@@

\begin{figure*}[!t]
\begin{center}
\includegraphics[width=6in]{waterfall3.png}
\end{center}
\caption{Different approaches to software development:  waterfall and agile (bottom left).}
\label{fig:waterfall}
\end{figure*}

@@

\section{Analysis of Phase Delay in 171 TSP Projects}

@@

\subsection{TSP Project Cost-to-fix curve}

Our first hypothesis was that defects found in later phases tend to require more time to fix than in earlier phases.

The distribution of defects found and fixed per phase in our data is shown in Figure~\ref{fig:fix-phase-dist}. A high percentage of defects were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections. (This distribution is similar to that observed for other projects that emphasized investment in software engineering quality assurance practices, for example the NASA robotics projects which had 58\% of defects removed before implementation began. \cite{me08a} Similar results are presented by Jones and Bonsignour, who show 52\% of pretest defects removed before entering implementation. \cite{jones12})

\todo{This paragraph is wrong now.} The median time to fix a defect in each phase is shown in \fig{fix-time-per-phase}. The median cost to fix a defect is low in the early phases, and then increases slightly during design and coding before peaking during the unit test phase. The median fix time then drops during the testing phases. Note that while the median time remains relatively consistent, the 75th percentile does increase during the testing phases.

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=3.3in]{fix-phase-dist.pdf}
\end{center}
\caption{Distribution of defects found and fixed by phase.}
\label{fig:fix-phase-dist}
\end{figure}

\begin{figure*}[!ht]

 \renewcommand{\baselinestretch}{0.7}
 \scriptsize
\begin{center}
\begin{tabular}{r|rr|rl}
Phase & samples & total time&\multicolumn{2}{l}{Total time w.r.t. first phase}\\\hline
~\\
Planning&229&12625&1.00&*\\
~\\
Reqts&590&24182&1.91&**\\
ReqtsInspect&2533&92723&7.34&********\\
ReqtsReview&1762&67471&5.34&******\\
~\\
HLD&78&6856&0.54&*\\
HLDReview&472&12821&1.01&**\\
HLDInspect&1366&35291&2.79&***\\
~\\
Design&620&21573&1.70&**\\
DesignInspect&8373&229937&18.21&*******************\\
DesignReview&3043&92823&7.35&********\\
~\\
Code&3029&220058&17.43&******************\\
CodeReview&4774&163091&12.91&*************\\
CodeInspect&11406&334442&26.49&***************************\\
~\\
UnitTest&9177&352245&2.790&****************************\\
IntTest&2761&160559&12.71&*************\\
SysTest&2059&105796&8.37&*********\\
AcceptTest&1045&42717&0.338&****\\
\end{tabular}
\end{center}
\caption{Times per phase (minutes).}
\label{fig:fix-time-per-phase}
\end{figure*}

For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}. Cost-to-fix for TSP is total time spent in each review, inspection, or test phase (reported in \fig{fix-time-per-phase}) / \# of defects found in phase (reported in \fig{raw}). It is clear to see that the trend is relatively flat, closer to the pattern exhibited by the Royce case study (which also focused on early defect removal) than the exponential growth curve.

\begin{figure}[!ht]
 \includegraphics[width=3.3in]{boehm-overlay.pdf}
 \caption{Cost-to-fix curve from \fig{cost-to-fix} overlayed with case study from~\cite{Royce98} and TSP data. Scale is relative to the phase with the lowest cost to fix. }
 \label{fig:cost-to-fix-tsp}
 \end{figure}

Given the above evidence, we reject $H_1$: The cost-to-fix of defects monotonically increases across phases.

@@

Our second hypothesis was that defects that exhibit higher phase delay (i.e. that persist in the software over more lifecycle phases) tend to require more effort to fix than those with less phase delay. To investigate this, we calculate phase delay as simply the number of project phases (the set used in our data set is shown in \fig{waterfall}) during which the defect existed in the system, i.e. the ordinal rank of removal phase minus the ordinal rank of injection phase. Sub-phases (such as review and inspection) are included in this calculation.

In \fig{raw}, we present the median fix times of defects based on when they were injected and ultimately removed. Defects injected Before Development (labelled "B4Dev" in \fig{waterfall}) follow the most severe scale up as development phases pass, but not on the exponential scale traditionally seen (\fig{cost-to-fix}). Defects injected in Requirements, Design, and Coding phases exhibit virtually no scale-up regardless of the phase in which they are fixed.

Given the above evidence, we reject $H_2$: the cost-to-fix of defects monotonically increases as the phase delay increases.

\begin{figure*}[!t] 
 \renewcommand{\baselinestretch}{0.7}
 \scriptsize
\begin{center}
\begin{tabular}{r|rrr|ll|rl}
  Sample&\multicolumn{3}{c|}{Percentiles}\\ 
size & 25th & 50th & 75th & Phase injected & Phase removed & \multicolumn{2}{l}{Scale up w.r.t. to first phase}\\\hline
\\
 48&   14&   30&   51&Before&DesignInspect &1.00 & **********  \\
 94&   22&   45&   80&Before&CodeInspect &1.50 & ***************  \\
 53&   13&   28&   49&Before&CodeReview &0.93 & **********  \\
153&   38&   77&  130&Before&UnitTest &2.56 & **************************  \\
127&   30&   65&  128&Before&IntTest &2.16 & **********************  \\
 92&   24&   56&   89&Before&SysTest &1.86 & *******************  \\

\\
189&   46&   93&  155&Reqts&ReqtsInspect &1.00 & **********  \\
184&   45&   92&  155&Reqts&ReqtsReview &0.98 & **********  \\
102&   24&   49&   83&Reqts&DesignInspect &0.52 & ******  \\
 61&   14&   33&   55&Reqts&DesignReview &0.35 & ****  \\
 59&   18&   38&   60&Reqts&CodeInspect &0.40 & ****  \\
 91&   22&   49&  104&Reqts&UnitTest &0.52 & ******  \\
100&   33&   67&  120&Reqts&IntTest &0.72 & ********  \\
 78&   28&  122&  237&Reqts&SysTest &1.31 & **************  \\
\\
207&   50&  101&  156&Design&DesignInspect &1.00 & **********  \\
210&   51&  103&  166&Design&DesignReview &1.01 & ***********  \\
163&   39&   79&  135&Design&CodeInspect &0.78 & ********  \\
134&   32&   66&  116&Design&CodeReview &0.65 & *******  \\
238&   58&  119&  183&Design&UnitTest &1.17 & ************  \\
172&   42&   88&  158&Design&IntTest &0.87 & *********  \\
163&   40&   80&  161&Design&SysTest &0.79 & ********  \\
 80&   21&   45&   73&Design&AcceptTest &0.44 & *****  \\

\\
130&   31&   65&  122&Code&AcceptTest &1.00 & **********  \\
228&   56&  113&  176&Code&CodeInspect &1.73 & ******************  \\
235&   57&  115&  177&Code&CodeReview &1.76 & ******************  \\
243&   59&  125&  206&Code&IntTest &1.92 & ********************  \\
191&   46&   95&  173&Code&SysTest &1.46 & ***************  \\
269&   66&  133&  206&Code&UnitTest &2.04 & *********************  \\
 \end{tabular}
\end{center}
\caption{Distribution of fix times seen in SEI TSP data.}
\label{fig:raw}
\end{figure*}

@@

\emph{Phase delay}, as depicted in the scale up column in \fig{cost-to-fix-tsp} \fig{raw}, is a discrete value that does not consider the total time that a defect is latent in the system. Thus, the phase delay of two defects injected in Code but found in System Test will be the same even though one was found at the beginning of test while the other at the end of a potentially lengthy test period. Ideally, one could examine the total real time a defect was in a system from the point at which is was injected to the point at which it was discovered. It may be possible to gather such information if the project keeps rigorous change logs for its requirements, design, and code artifacts.

\subsection{Internal validity}
Confounding variables are a threat to any study. In this research, we rely on the substantial sample size of 171 TSP projects and over 47,000 defect logs to avoid confounding effects that may arise from the nature of the projects, teams, or defects. Similarly, the 10-year data collection period should help to ameliorate any maturation or history effects.  As described in \S\ref{sec:data-collection}, all TSP teams are required to contribute time and defect data to the SEI, and thus there should be no selection bias in this sample compared to the overall population of TSP projects. However, there is likely selection bias in the teams that elect to use TSP compared to the entire population of software development teams.

>>>>

\conferenceinfo{FSE}{'15 Bergamo, Italy}
\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase  Fix Time?}
\numberofauthors{5}
\author{
\alignauthor 
Tim Menzies \\
       \affaddr{Dept. of Computer Science}\\
       \affaddr{North Carolina State U, USA}\\
       \email{tim.menzies@gmail.com}
\alignauthor 
William Nichols \\
        \affaddr{Software Engineering Institute}\\
        \affaddr{Carnegie Mellon U, USA}
        \email{wrn@sei.cmu.edu} 
\alignauthor 
Lucas Layman \\
       \affaddr{Fraunhofer CESE} \\
       \affaddr{College Park, USA}\\ 
       \email{llayman@fc-md.umd.edu}
\and
\alignauthor 
Forrest Shull \\
        \affaddr{Software Engineering Institute}\\
        \affaddr{Carnegie Mellon U, USA}
        \email{fjshull@sei.cmu.edu} 
\alignauthor 
Carter Pape \\
       \affaddr{Dept. of Computer Science}\\
       \affaddr{North Carolina State U, USA}\\
       \email{carterpape@gmail.com}
} 
\maketitle
\begin{abstract}
Does
repair time increase dramatically
the longer an issue persists in a system?
Many industrial practitioners and academics this   to be true.
This   belief is used to justify 
major investments in  software development
 processes and methods.

This
paper reports on the largest study on phase delay yet published.
Based on a sample of 
171 software projects from around the world (2006 and 2014),
 we find that this belief is  
incorrect. Specifically: the    time  to resolve an
issue   does {\em not} increase
the longer the issue remains in the system.

This result raises the question: how do widely-held beliefs like phase delay survive so long  without  critical reassessment?  As a community, we need to consider how (and when) we reflect on our prominent hypotheses.
\end{abstract}

@@

\section{Introduction}
In 2013-2014, 
eleven  million programmers~\cite{pettey14} and
half a trillion dollars~\cite{avram14} was spent on information technology.
Such a large and growing effort should be managed and optimized via  well-researched conclusions.

It is standard practice
in other fields such as medicine,
to continually revisit old conclusions~\cite{prasad13}.
Accordingly, this paper revisits
the {\em phase delay effect}
i.e.,  the rule of thumb stating that the time required to resolve an issue in a software project increases dramatically 
the longer it remains in the system. Basili \& Boehm comment that since the 1980s,  phase delay:
\begin{quote}
...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes~\cite{boehm01}.
\end{quote}
Basili and Boehm are not alone in this assessment of the significance of the phase delay effect. Our surveys of industrial practitioners (shown in \S\ref{survey}) reveal the phase delay effect to be
one of the most strongly-held beliefs amongst commercial software engineers.
Previously~\cite{me08a}, we used phase delay to justify the software practice
of 
Independent Verification \& Validation (IVV):
\bi
\item
IVV has been criticized as
ineffective. IVV engineers are external reviewers, so may be unfamiliar with the project details.
Hence,    IVV   can find the same or fewer bugs
as  developers.
\item
On the other hand,   when the IVV team finds bugs earlier than the developers then, due to the phase delay effect, this provides overall cost savings to the project. 
\ei
Note that this argument critically depends on the assumption that   fixing later-phase issues is  significantly more expensive than fixing them earlier (as does the views of Boehm and Basili and all the industrial practitioners
we interviewed in \S\ref{survey}). Accordingly, in this paper, we investigate several hypotheses:
\bi
    \item $H_1$: The cost to resolve issues dramatically increases across phases.
    \item $H_2$: The cost to resolve issues dramatically increases as the phase delay increases.
\ei
As we explored the above, we found a related hypothesis was believed by both professional software engineers and some academics:
\bi
 \item $H_3$: Requirements errors are the most expensive to fix.
\ei
The rest of the paper investigates these  hypotheses. First, we conduct a survey of practitioners and researchers to gauge their belief in the phase delay effect. Second, we  analyze 171 software  projects developed in the period 2006 to 2014.

% The conclusions of this paper is that it is erroneous to believe in phase delay.
% When we looked for supporting evidence for that effect, we found that:
% \be 
% \item The evidence for that effect in the historical literature is very weak (to say the least);
% \item  In  a large sample of   contemporary software projects, there is no evidence for phase delay.
%  \ee
% The rest of this paper present evidence for the above two points. 
% The first point is  documented via  a literature survey and the second point is demonstrated
% via an analysis of

@@

The core of this paper's arguments are two figures: Figure~\ref{fig:fix-phase-dist} and  \fig{raw}.
Lest the reader find this too little  to justify our conclusions, we note that it took months
of work to distill this data from our 171 projects, after which the five authors spent weeks
exploring and debating the nuances. To the best of our knowledge, this is the largest study on phase delay yet published (and it contains data from orders of magnitude
more projects than used in prior work~\cite{Fagan76,Boehm76,Daly77,Stephenson76,Shull02,Royce98,Clutterbuck09,Elssamadisy02}).
To ensure reproducibility and unlike previous work,
all the data  used in this study is available in the PROMISE
repository (openscience.us/repo).

@@

Belief in general
principles of software
engineering are common to both research and practice. Professional societies assume such generalities exist when they offer
 lists of supposedly general ``best practices'' such as
the IEEE 1012 standard for software verification~\cite{1012}.  Endres \& Rombach's comment that 
\begin{quote}
Based on repeated and consistent observations, key lessons of these fields can now be formulated into rules or even laws, providing initial building blocks towards a theoretical foundation that is essential for further research, for teaching and for the practice of software development~\cite{endres03,rombach11}
\end{quote}
 Endres \& Rombach offer dozens of lessons of software engineering~\cite{rombach11}.
 Many other 
widely-cited researchers  do the same; e.g.
Glass~\cite{glass02}; Jones~\cite{jones07}; Boehm~\cite{boehm00b}.
Budgen \& Kitchenham seek to reorganize SE research using
general
conclusions drawn from a larger number of studies~\cite{kitch04,budgen09}.
Nevertheless, there are many empirical  findings that 
cast doubt on the existence of  ``key lessons  that can be formulated
as general laws of SE'':
\bi
\item
Turhan~\cite{me12d} lists 28 studies with contradictory conclusions
on the relation of OO measures to defects.  Those results
 directly  contradict some of the laws listed by 
Endres \& Rombach~\cite{endres03}.

@@

The cognitive issues that complicate medicine are also found in software engineering.
According to Passos et al.~\cite{passos11}, commercial developers
are all too willing to believe in general
truisms.
They  warn that developers
usually assume that the truisms they learn from a few past
projects are general to 
all their future projects:
\begin{quote}\label{q:pass}
...the past experiences were taken into account without 
much consideration for their context~\cite{passos11}.  
\end{quote}
The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} concur with Passos et al. They report that 
  supposed software engineering    ``experts'' rarely use lessons
  from past projects to improve their future reasoning~\cite{jorgensen09}. 
 They note that
when the experts
  fail to revise their beliefs, this leads to poor
 conclusions and software projects  (see examples in~\cite{jorgensen09}).
%% This leads to  detrimental
%% situations where groups or 
%% individuals working on the same project have conflicting beliefs, which they never debate
%or refine or unify.

@@

That said,  the phase delay myth persists. Hence, this paper.

@@

\subsection{Survey:  Phase Delay and Cost-to-fix Belief}\label{survey} 
Yet another reason to study the phase delay effect is that it is a dominant belief in industrial SE.
To test whether this was indeed true, we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
understanding of SE theory, then the phase delay effect may well influence their judgment.

@@

@@

@@

\section{Analysis of Phase Delay in 171 TSP Projects}
Our units of analysis are:
\bi
    \item \emph{defects} - individual defects are recorded as line items in the defect logs uploaded to the SEMPR at the SEI. One or more defects are reported against a single \emph{plan item} in the time tracking logs, e.g., a review session, an inspection meeting, a test execution.
    \item \emph{time} - Time is tracked per person per plan item in the time-tracking logs, e.g. a 30 minute design review session involving 3 people will have three time log entries summing to 90 minutes. Time includes the time to analyze, repair, and validate a defect fix.
    \item \emph{time per defect} - The total \# of defects found in a plan item during a removal phase divided by the total time spent on that plan item in that phase.
\ei

@@

\begin{figure}[!t]

%\hspace{-0.7cm}
\begin{center}
\includegraphics[height=2in]{fix-phase-dist.pdf}
\end{center}
%~\\

\caption{Distribution of defects found and fixed by phase.}
\label{fig:fix-phase-dist}
\end{figure}

\begin{figure}[!t] 
%\vspace{-1cm}
%\hspace{-0.35in}
\begin{center}
\includegraphics[width=3in]{boehm-overlay.pdf}
 \end{center}
% ~\\~\\
%~\\~\\~\\

 \caption{Cost-to-fix curve from \protect\fig{cost-to-fix} 
 overlayed with case study from~\cite{Royce98} and TSP data. Scale is relative to the phase with the lowest cost to fix. }\label{fig:cost-to-fix-tsp}
 \end{figure}

\begin{figure*}[!t] 
 \renewcommand{\baselinestretch}{0.7}
 \scriptsize
\begin{center}
\begin{tabular}{r|rrr|ll|rl}  % I attach your cursor. thrust, parry, yo!
  Sample&\multicolumn{3}{c|}{Percentiles}\\ 
size & 25th & 50th & 75th & Phase injected & Phase removed & \multicolumn{2}{l}{Scale up w.r.t. to first phase}\\\hline
\\
 48&   14&   30&   51&Before&DesignInspect &1.00 & **********  \\
 53&   13&   28&   49&Before&CodeReview &0.93 & **********  \\
 94&   22&   45&   80&Before&CodeInspect &1.50 & ***************  \\
153&   38&   77&  130&Before&UnitTest &2.56 & **************************  \\
127&   30&   65&  128&Before&IntTest &2.16 & **********************  \\
 92&   24&   56&   89&Before&SysTest &1.86 & *******************  \\

\\
184&   45&   92&  155&Reqts&ReqtsReview &1.00 & **********  \\
189&   46&   93&  155&Reqts&ReqtsInspect &1.00 & **********  \\
 61&   14&   33&   55&Reqts&DesignReview &0.35 & ****  \\
102&   24&   49&   83&Reqts&DesignInspect &0.52 & ******  \\
 59&   18&   38&   60&Reqts&CodeInspect &0.40 & ****  \\
 91&   22&   49&  104&Reqts&UnitTest &0.52 & ******  \\
100&   33&   67&  120&Reqts&IntTest &0.72 & ********  \\
 78&   28&  122&  237&Reqts&SysTest &1.31 & **************  \\
\\
210&   51&  103&  166&Design&DesignReview &1.01 & ***********  \\
207&   50&  101&  156&Design&DesignInspect &1.00 & **********  \\
134&   32&   66&  116&Design&CodeReview &0.65 & *******  \\
163&   39&   79&  135&Design&CodeInspect &0.78 & ********  \\
238&   58&  119&  183&Design&UnitTest &1.17 & ************  \\
172&   42&   88&  158&Design&IntTest &0.87 & *********  \\
163&   40&   80&  161&Design&SysTest &0.79 & ********  \\
 80&   21&   45&   73&Design&AcceptTest &0.44 & *****  \\

\\
235&   57&  115&  177&Code&CodeReview &1.00 &  ******************  \\
228&   56&  113&  176&Code&CodeInspect &0.99 & ******************  \\
243&   59&  125&  206&Code&IntTest &1.08 &     *******************  \\
191&   46&   95&  173&Code&SysTest &0.82 & ******** \\
269&   66&  133&  206&Code&UnitTest &1.15 & ************  \\
130&   31&   65&  122&Code&AcceptTest &0.56 & ******  \\
 \end{tabular}
\end{center}
\caption{Distribution of fix times seen in SEI TSP data. Left hand side bars
show scale up on median values.}
\label{fig:raw}
\end{figure*}

\subsection{TSP Project Cost-to-fix curve}
This section discusses our first hypothesis ($H_1$) that   that  the cost-to-resolve issues dramatically increases across phases.

The distribution of defects found and fixed per phase in our data is shown in Figure~\ref{fig:fix-phase-dist}. A high percentage of defects (44\%) were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections. (This distribution is similar to that observed for other projects that emphasized investment in software engineering quality assurance practices. For example, Jones and Bonsignour report 52\% of pretest defects removed before entering implementation, for large projects that focus on upfront defect removal techniques \cite{jones12}. NASA robotics projects had a slightly higher percentage (58\%) of defects removed before implementation began, although these had invested in IV\&V on top of other forms of defect removal \cite{me08a}.)

The first thing to note in Figure~\ref{fig:fix-phase-dist}, 
is that this data does not follow the exponential increase feared by Beck's \fig{curve1}:
\bi
\item There is much ``up and down'' in the chart. 
\item If we separate out just the {\em Inspect} or {\em Review} data, we can see that there is some increase in {\em Design} and {\em Code}. However, that effect could be a linear or be just noise. In either case,
it does not look like \fig{curve1}.
\item The time in final testing is particularly low suggesting that few defects survived into testing
(and hence, there were few instances where code required massive rework).
\ei

For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}. Cost-to-fix for TSP is total time spent in each review, inspection, or test-related phase / \# of defects found in phase (reported in \fig{raw}). It is clear to see that the trend is relatively flat, closer to the pattern exhibited by the Royce case study (which also focused on early defect removal) than the exponential growth curve.

%\begin{table}[ht]
%\renewcommand{\baselinestretch}{0.7}
% \scriptsize
%\centering
%\begin{tabular}{ll|ll}
%  Phase num & Name & \# defects & percentage \\ 
%  \hline
%1 & BeforeDevelopment &   0 & 0.00 \\ 
%  2 & Planning &   4 & 0.00 \\ 
%  3 & Reqts &  25 & 0.10 \\ 
%  4 & ReqtsReview & 2304 & 6.50 \\ 
%  5 & ReqtsInspect & 2630 & 7.40 \\ 
%  6 & HLD &   0 & 0.00 \\ 
%  7 & HLDReview & 790 & 2.20 \\ 
%  8 & HLDInspect & 818 & 2.30 \\ 
%  9 & Design &  94 & 0.30 \\ 
%  10 & DesignReview & 3668 & 10.30 \\ 
%  11 & DesignInspect & 5365 & 15.00 \\ 
%  12 & Code & 833 & 2.30 \\ 
%  13 & CodeReview & 6276 & 17.60 \\ 
%  14 & CodeInspect & 6623 & 18.60 \\ 
%  15 & UnitTest & 4283 & 12.00 \\ 
%  16 & IntTest & 1024 & 2.90 \\ 
%  17 & SysTest & 730 & 2.00 \\ 
%  18 & AcceptTest & 189 & 0.50 \\ 
% \end{tabular}
%\caption{Defects found per phase} 
%\label{fig:defect_per_phase}
%\end{table}

%\begin{figure}[!ht]
%
% \renewcommand{\baselinestretch}{0.7}
% \scriptsize
%\begin{center}
%\begin{tabular}{r|rr|rl}
%Phase & samples & total time&\multicolumn{2}{l}{Total time w.r.t. first phase}\\\hline
%~\\
%Planning&229&12625&1.00&*\\
%~\\
%Reqts&590&24182&1.91&**\\
%ReqtsInspect&2533&92723&7.34&********\\
%ReqtsReview&1762&67471&5.34&******\\
%~\\
%HLD&78&6856&0.54&*\\
%HLDReview&472&12821&1.01&**\\
%HLDInspect&1366&35291&2.79&***\\
%~\\
%Design&620&21573&1.70&**\\
%DesignInspect&8373&229937&18.21&*******************\\
%DesignReview&3043&92823&7.35&********\\
%~\\
%Code&3029&220058&17.43&******************\\
%CodeReview&4774&163091&12.91&*************\\
%CodeInspect&11406&334442&26.49&***************************\\
%~\\
%UnitTest&9177&352245&2.790&****************************\\
%IntTest&2761&160559&12.71&*************\\
%SysTest&2059&105796&8.37&*********\\
%AcceptTest&1045&42717&0.338&****\\
%\end{tabular}
%\end{center}
%\caption{Times per phase (minutes).}
%\label{fig:fix-time-per-phase}
%\end{figure}

Given the above evidence, we reject $H_1$: there is no clear evidence that  cost-to-fix of defects dramatically increases across phases.

@@

 \subsubsection{$H_2$: Cost-to-resolve Increases with Delay}

%we calculate phase delay as simply the number of project phases (the set used in our data set is shown in %\fig{waterfall}) during which the defect existed in the system, i.e. the ordinal rank of removal phase minus the %ordinal rank of injection phase. Sub-phases (such as review and inspection) are included in this calculation.
%
\fig{raw} shows the 25th, 50th, 75th percentile
of the time spent on {\em review} or {\em inspect} or {\em test} phases
in our 171 projects.
Note that, in TSP, when developers see issues, they enter {\em review} or 
{\em inspect} or {\em test}
until that issue is retired.

In that figure, the data is split out according to issues that were fixed in phase $Y$ after
being introduced in phase $X$. The data is sub-divided into tables according to $X$;
i.e. according to {\em before, planning, requirements, design} or {\em  code}.

The stars on the right-hand-side of that figure show the scale ups associated with the phase removal stages,
expressed as a ratio of the first item in that block (which, by definition, is 1.00 for
that first item).
If $H_2$ was correct, we would expect those bars in each sub-division to exhibit a  shape
like \fig{curve1}.

They do not. Hence, we reject $H_2$.

 \subsubsection{$H_3$: Requirements Repairs are Most Expensive}

As to $H_3$, the {\em requirements} the  median time spent fixing errors in {\em requirements} was 92,93  49,33,39,49,67,122 minutes\footnote{If this time seems long, then recall
that it includes the time required  to (a)~collect data and realize there is an error;
(b)~prepare a fix;  and (c)~apply some validation
procedure to check the fix (e.g. discuss it with a colleague or execute some tests).}.
This is:
\bi
\item
Clearly longer than median times seen {\em Before } and in {\em Planning}
(30,28,45,77,65,56,24,35,33);
\item
Somewhat shorter, but still not exponentially smaller,  than the median times see after {\em requirements} in {\em design}
(103,101, 79,66, 119,88, 80,45) and in {\em code} (65,115, 113,125, 95,133). 
\ei
For completeness, we ran an  
A12 test comparing the raw data (and not just the medians of \fig{raw}) and found
 a non-small effect only for the difference {\em requirements}
to {\em code}. But even though the effect was non-small, 
 the overall size of the difference
{\em requirements} to {\em code} was less than 30\% (i.e. much less than we would
expect from \fig{curve1}.

Based on the above, we reject   $H_3$.

@@

\emph{Phase delay}, as depicted in the scale up column in \fig{cost-to-fix-tsp} and \fig{raw}, is a discrete value that does not consider the total time that a defect is latent in the system. Thus, the phase delay of two defects injected in Code but found in System Test will be the same even though one was found at the beginning of test while the other at the end of a potentially lengthy test period. Ideally, one could examine the total real time a defect was in a system from the point at which is was injected to the point at which it was discovered. It may be possible to gather such information if the project keeps rigorous change logs for its requirements, design, and code artifacts.

\subsection{Internal validity}
Confounding variables are a threat to any study. In this research, we rely on the substantial sample size of 171 TSP projects and over 47,000 defect logs to avoid confounding effects that may arise from the nature of the projects, teams, or defects. Similarly, the 10-year data collection period should help to ameliorate any maturation or history effects.  As described in \S\ref{sec:data-collection}, all TSP teams are required to contribute time and defect data to the SEI, and thus there should be no selection bias in this sample compared to the overall population of TSP projects. However, there is likely selection bias in the teams that elect to use TSP compared to the entire population of software development teams. Further, we assume that each team has similar defect recording practices, and the TSP coaching provides guidance on what constitutes a defect. Nonetheless, individual developers and teams may apply their own internal rules for filtering defects, which would lead to inconsistent reporting thresholds among the projects in our sample.
